---
title: "Chapter 1 - Introduction to Data"
author: 'Samuel I Kigamba'
output:
  html_document:
    df_print: paged
  pdf_document:
    extra_dependencies:
    - geometry
    - multicol
    - multirow
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Smoking habits of UK residents**. (1.10, p. 20) A survey was conducted to study the smoking habits of UK residents. Below is a data matrix displaying a portion of the data collected in this survey. Note that "$\pounds$" stands for British Pounds Sterling, "cig" stands for cigarettes, and "N/A" refers to a missing component of the data.

\begin{center}
\scriptsize{
\begin{tabular}{rccccccc}
\hline
	& sex 	 & age 	& marital 	& grossIncome 					     & smoke & amtWeekends	& amtWeekdays \\ 
\hline
1 	& Female & 42 	& Single 	& Under $\pounds$2,600 			     & Yes 	 & 12 cig/day   & 12 cig/day \\ 
2 	& Male	 & 44	& Single 	& $\pounds$10,400 to $\pounds$15,600 & No	 & N/A 			& N/A \\ 
3 	& Male 	 & 53 	& Married   & Above $\pounds$36,400 		     & Yes 	 & 6 cig/day 	& 6 cig/day \\ 
\vdots & \vdots & \vdots & \vdots & \vdots 				             & \vdots & \vdots 	    & \vdots \\ 
1691 & Male  & 40   & Single 	& $\pounds$2,600 to $\pounds$5,200   & Yes 	 & 8 cig/day 	& 8 cig/day \\   
\hline
\end{tabular}
}
\end{center}


(a) What does each row of the data matrix represent?
        
        Each row repsents an observational unit, or a single study subject in this case, a UK resident.
        
(b) How many participants were included in the survey?
        
        Total number of participants is 1691 uk residents. This is our sample.
        
(c) Indicate whether each variable in the study is numerical or categorical. If numerical, identify as continuous or discrete. If categorical, indicate if the variable is ordinal.
        
        Gender: Categorical, nominal
        Age: Numerical, Discrete
        Marital Status: Categorical, Ordinal
        Highest Qualification: Categorical, nominal
        Nationality: Categorical, nominal
        Ethnicity: Categorical, Ordinal
        Gross Income: categorical, Ordinal
        Region: Categorical, nominal
        Smoke: Categorical, nominal
        amtWeekends: Numerical, Discrete
        amtWeedays: Numerical, Discrete
        Type: Categorical, nominal

```{r, eval=FALSE, echo=FALSE}
# NOTE: It is not absolutely necessary to load the data to answer these 
# questions, but is here for further exploration. 
if(!file.exists('11263-Smoking_tcm86-13253.xls')) {
	download.file('https://github.com/jbryer/DATA606Fall2019/blob/master/course_data/11263-Smoking_tcm86-13253.xls?raw=true', 
				  destfile = '11263-Smoking_tcm86-13253.xls')
}
smoking <- readxl::read_excel('11263-Smoking_tcm86-13253.xls')
```





--------------------------------------------------------------------------------

\clearpage

**Cheaters, scope of inference**. (1.14, p. 29) Exercise 1.5 introduces a study where researchers studying the relationship between honesty, age, and self-control conducted an experiment on 160 children between the ages of 5 and 15[^1]. The researchers asked each child to toss a fair coin in private and to record the outcome (white or black) on a paper sheet, and said they would only reward children who report white. Half the students were explicitly told not to cheat and the others were not given any explicit instructions. Differences were observed in the cheating rates in the instruction and no instruction groups, as well as some differences across children창????s characteristics within each group.

(a) Identify the population of interest and the sample in this study.
        
        Population of interest is all children between the ages of 5 and 15.
        Sample is 160 children between the ages of 5 and 15.
        
(b) Comment on whether or not the results of the study can be generalized to the population, and
if the findings of the study can be used to establish causal relationships.
        
        This study is experimental, and as such can be generalized to the general children population, however its only applicable to those between the ages of 5 and 15.
        To establish a causal relationship there is need for additional work in the area of reward, to see if cheating patters vary with the reward, or is general.
        

[^1]: Alessandro Bucciol and Marco Piovesan. "Luck or cheating? A field experiment on honesty with children". In: Journal of Economic Psychology 32.1 (2011), pp. 73창????78. Available at https://papers.ssrn.com/sol3/papers.cfm?abstract_id=1307694



--------------------------------------------------------------------------------

\clearpage

**Reading the paper**. (1.28, p. 31) Below are excerpts from two articles published in the NY Times:

(a) An article titled Risks: Smokers Found More Prone to Dementia states the following:

"Researchers analyzed data from 23,123 health plan members who participated in a voluntary exam and health behavior survey from 1978 to 1985, when they were 50-60 years old. 23 years later, about 25% of the group had dementia, including 1,136 with Alzheimer's disease and 416 with vascular dementia. After adjusting for other factors, the researchers concluded that pack-a- day smokers were 37% more likely than nonsmokers to develop dementia, and the risks went up with increased smoking; 44% for one to two packs a day; and twice the risk for more than two packs."

Based on this study, can we conclude that smoking causes dementia later in life? Explain your reasoning.
        
        Concluding that smoking causes dimentia later in life would amount to generalizing the results of the study, which is not possible with observational studies. For causal connection to be established an experiment must be conducted. Further more the study group was made up of volunteers and might potray elements of bias in as far as the respondents are concerned.

(b) Another article titled The School Bully Is Sleepy states the following:

"The University of Michigan study, collected survey data from parents on each child's sleep habits and asked both parents and teachers to assess behavioral concerns. About a third of the students studied were identified by parents or teachers as having problems with disruptive behavior or bullying. The researchers found that children who had behavioral issues and those who were identified as bullies were twice as likely to have shown symptoms of sleep disorders."

A friend of yours who read the article says, "The study shows that sleep disorders lead to bullying in school children." Is this statement justified? If not, how best can you describe the conclusion that can be drawn from this study?
        
        The statement is not justifiable, from the study there is an association between sleep disorder and bullying but there is no evidence that either leads to the other. There is need to assess other variables to determine this relationship.
        
        





--------------------------------------------------------------------------------

\clearpage

**Exercise and mental health.** (1.34, p. 35) A researcher is interested in the effects of exercise on mental health and he proposes the following study: Use stratified random sampling to ensure rep- resentative proportions of 18-30, 31-40 and 41-55 year olds from the population. Next, randomly assign half the subjects from each age group to exercise twice a week, and instruct the rest not to exercise. Conduct a mental health exam at the beginning and at the end of the study, and compare the results.

(a) What type of study is this?
        
        This is an experiment, specifically a blocked randomized experiment.
        
(b) What are the treatment and control groups in this study?
        
        The treatment group is the half that was instructed to excercise twice a week, while the control group is the second half that was instructed not to excercise.
        
(c) Does this study make use of blocking? If so, what is the blocking variable?
        
        Yes, the blocking variable is age(18-3-, 31-40 and 41-55 year olds).
        
(d) Does this study make use of blinding?
        
        There is no blinding since both the administrators and the experiments groups know their roles. There is no placebo used.
        
(e) Comment on whether or not the results of the study can be used to establish a causal rela- tionship between exercise and mental health, and indicate whether or not the conclusions can be generalized to the population at large.
        
        The experiment design seems to be sufficient, being a randomized blocked experiment, however the control group might give misleading results due to the fact that thair response might be as a result of a deviation from routine. There response to mental health might be as a result of not excercising, which they are used to. Selection of a control group made up of people who are already not excercising and thus no change in their routine, might give more accurate results.
        
(f) Suppose you are given the task of determining if this proposed study should get funding. Would you have any reservations about the study proposal?
        
        Yes, the control group would need to be reconfigured to provide accurate results.
        
        



